The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-02-24 src/ms.tex
Fix latex error
continue draft section on interoperability
Signed-off-by: nstarman <nstarkman@protonmail.com>

\newcommand{\astropySpectralCoord}{\astropyapidoc{coordinates.SpectralCoord}{SpectralCoord}}
\newcommand{\astropySkyCoord}{\astropyapidoc{coordinates.SkyCoord}{SkyCoord}}
\newcommand{\astropyICRS}{\astropyapidoc{coordinates.builtin_frames.ICRS}{ICRS}}
\newcommand{\astropyGalacticLSR}{\astropyapidoc{coordinates.builtin_frames.
GalacticLSR}{GalacticLSR}}
\newcommand{\astropyAltAz}{\astropyapidoc{coordinates.builtin_frames.AltAz}{AltAz}}
\newcommand{\astropyGalactocentric}{\astropyapidoc{coordinates.builtin_frames.Galactocentric}{Galactocentric}}
\newcommand{\astropyCosmology}{\astropyapidoc{cosmology.Cosmology}{Cosmology}}
\newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
\newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
\newcommand{\astropyQTable}{\astropyapidoc{table.QTable}{QTable}}
\newcommand{\astropyTime}{\astropyapidoc{time.Time}{Time}}
\newcommand{\astropyLeapSeconds}{\astropyapidoc{time.LeapSeconds}{LeapSeconds}}
\newcommand{\astropyDistribution}{\astropyapidoc{uncertainty.Distribution}{Distribution}}
\newcommand{\astropyUnit}{\astropyapidoc{units.Unit}{Unit}}
\newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
\newcommand{\astropyMasked}{\astropyapidoc{utils.masked.Masked}{Masked}}
\newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}

@@

  % New features capture a lot of attention. While exciting, the arguably more
  % significant improvements are those of interoperability -- ensuring existing
  % features connect smoothly together.

  % MHvK: this repeats the same thing 3 times; try to use just one sentence!
  The modular nature of \astropypkg means that not all sub-packages deeply
  integrate with one another -- and similarly between separate \python
  libraries. Because \astropy is situated at the nexus between scientific
  computing into astronomy, interoperability has been and continues to be an
  important focus for \astropy, both within the various \astropypkg sub-packages
  and with other important \python libraries. While increasing inter and
  intra-package integration will be a perpetual goal -- with feature additions
  in one place requiring updates in another -- \astropy has made impactful
  progress with units, tables, and I/O.
  % Better list? Also better transition
  If done right, all a user notices is that code using many different python
  features and libraries ``just works''.  Here, we describe a few
  particular efforts towards this goal.

@@

    \astropyQuantity\ is a backbone of \astropy, leveraging the power of
    \package{numpy} and adding units. Previously, many \package{numpy} functions
    would strip a \astropyQuantity\ of its units (or fail outright), limiting
    \astropyQuantity's usefulness. Now, advances in \package{numpy} function
    overloading (see for example
    \href{https://numpy.org/neps/nep-0018-array-function-protocol.html}{NEP 18})
    mean \astropyQuantity\ works with almost all \package{numpy} (version 1.17+)
    functions.
    % MHvK: I don't think the following is true, so commented it out.
    % These advances have similarly benefitted many
    % \package{numpy}-dependent packages, like \package{scipy} and their sparse
    % matrices. And again \astropypkg benefits since more \package{scipy}
    % functions now work with \astropyQuantity.

    In the remaining gaps, the \package{numpy} and \astropyQuantity\
    interoperability efforts are ongoing. For some functions, such as in
    \href{https://numpy.org/doc/stable/user/basics.rec.html#module-numpy.lib.recfunctions}{\package{numpy}'s
    module} for manipulating structured arrays, compatibility only requires
    exending the existing \package{numpy}-\astropypkg bridge framework.
    Community interest, in the form of a Feature Request (or better yet Pull
    Request) would be sufficient to see this realized. For a few remaining functions,
    discussed further in the
    \href{https://docs.astropy.org/en/stable/known_issues.html#known-deficiencies}{\astropy
    documentation}, the \package{numpy} framework does not yet allow for full
    interoperability with \astropy. A goal of \astropy, and the scientific
    \python community is to enhance and implement the frameworks, allowing
    \astropyQuantity\ to propagate units seamlessly across the whole ecosystem of
    \package{numpy}-like projects.

  \paragraph{Astropy in Time and on Units}

    % Bad sentence ↓. Rewrite.
    Not only did \astropyunits gain powerful integrations with \package{numpy},
    there were many intra-package improvements.

    \astropymodeling can add units to otherwise unit-less models.

    A new module –
    \astropycosmologyunits
    – has been added to the cosmology sub-package for defining and collecting
    cosmological units and equivalencies. The unit \texttt{littleh} and
    equivalency
    \astropyapidoc{cosmology.units.with_H0}{with\_H0}
    are deprecated from the main \astropyunits sub-package and moved to
    \astropycosmologyunits. A new unit, \texttt{redshift}, is added for tracking
    factors of cosmological redshift.
    % As this is a pseudo-unit an equivalency
    % \astropyapidoc{cosmology.units.dimensionless_redshift}{dimensionless\_redshift}
    % is added (and enabled by default) to allow for redshift - dimensionless
    % conversions.
    To convert between redshift and other cosmological distance
    measures, e.g. CMB temperature or comoving distance, the equivalency
    \astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
    is also added. This equivalency is actually a composite of other
    equivalencies,
    % : \astropyapidoc{cosmology.units.redshift_distance}{redshift\_distance},
    % \astropyapidoc{cosmology.units.redshift_hubble}{redshift\_hubble}, and
    % \astropyapidoc{cosmology.units.redshift_temperature}{redshift\_temperature},
    which may be used separately.

    % IN PROGRESS (ADAPTING FROM THE WHATSNEW).
    The FITS Time standard paper defines the formats and keywords used to
    represent timing information in FITS files. The Astropy FITS package now
    provides support for reading and writing native Time table columns using
    this format. This is described within the FITS unified I/O interface and
    examples of usage can be found in the Astropy native objects (mixin columns)
    section.

@@

@@

    First, the class-based architecture allows for the creation of custom
    registries. An application of this is in \astropycosmology, which
    has two different \textit{kinds} of registries for \astropyCosmology: one
    for reading and writing files, and another for converting between python
    objects.

    Second, \textit{any} kind of object is now compatible with
    \astropyiosubpkg{registry}. Using \astropyiosubpkg{registry} for custom
    classes means a user does not have to create a whole new I/O system and can
    offer a unified and familiar API.

    % TODO Better transition sentence
    The imrovements to \astropy's I/O architecture are mirrored in the increased support for I/O options.

@@

  An important focus of \astropy has been to increase the precision and accuracy
  of its functionality, while, where appropriate, making sure results obtained
  using older versions of \astropypkg are reproducible.
  % Something about ScienceState
  Notable improvements have been made to \astropytime, \astropyconstants,
  \astropycoordinates, and \astropycosmology.

  \paragraph{Time}

    In astronomy, time accuracy down to the second is frequently important.
    Whether for planning observations or crunching data, missing seconds can
    meaningfully impact results. Before, when using \astropyTime in \astropytime
    leap seconds had to be manually applied; now leap seconds are applied
    automatically. Moreover, \astropypkg updates internal time data-files to
    ensure that the correct leap second adjustment is always used. For
    reproducibility, e.g. with an old code, \astropyLeapSeconds may also be
    manually applied.

>>>>

\newcommand{\astropySpectralCoord}{\astropyapidoc{coordinates.SpectralCoord}{SpectralCoord}}
\newcommand{\astropySkyCoord}{\astropyapidoc{coordinates.SkyCoord}{SkyCoord}}
\newcommand{\astropyICRS}{\astropyapidoc{coordinates.builtin_frames.ICRS}{ICRS}}
\newcommand{\astropyGalacticLSR}{\astropyapidoc{coordinates.builtin_frames.
GalacticLSR}{GalacticLSR}}
\newcommand{\astropyAltAz}{\astropyapidoc{coordinates.builtin_frames.AltAz}{AltAz}}
\newcommand{\astropyGalactocentric}{\astropyapidoc{coordinates.builtin_frames.Galactocentric}{Galactocentric}}
\newcommand{\astropyCosmology}{\astropyapidoc{cosmology.Cosmology}{Cosmology}}
\newcommand{\astropyFlatLambdaCDM}{\astropyapidoc{cosmology.FlatLambdaCDM}{FlatLambdaCDM}}
\newcommand{\astropyModel}{\astropyapidoc{modeling.Model}{Model}}
\newcommand{\astropyTable}{\astropyapidoc{table.Table}{Table}}
\newcommand{\astropyQTable}{\astropyapidoc{table.QTable}{QTable}}
\newcommand{\astropyTime}{\astropyapidoc{time.Time}{Time}}
\newcommand{\astropyLeapSeconds}{\astropyapidoc{time.LeapSeconds}{LeapSeconds}}
\newcommand{\astropyDistribution}{\astropyapidoc{uncertainty.Distribution}{Distribution}}
\newcommand{\astropyUnit}{\astropyapidoc{units.Unit}{Unit}}
\newcommand{\astropyQuantity}{\astropyapidoc{units.Quantity}{Quantity}}
\newcommand{\astropyMasked}{\astropyapidoc{utils.masked.Masked}{Masked}}
\newcommand{\astropyScienceState}{\astropyapidoc{utils.state.ScienceState}{ScienceState}}

@@

  Because \astropy is modular and situated at the nexus between scientific
  computing and astronomy, interoperability is an important focus for \astropy,
  both within the various \astropypkg sub-packages and with other important
  \python libraries. If done right, all a user notices is that code using many
  different python features and libraries ``just works''. Here, we describe a
  few particular efforts towards this goal.

@@

    \astropyQuantity\ is a backbone of \astropy, leveraging the power of
    \package{numpy} and adding units. Previously, many \package{numpy} functions
    would strip a \astropyQuantity\ of its units (or fail outright), limiting
    \astropyQuantity's usefulness. Now, advances in \package{numpy} function
    overloading (see for example
    \href{https://numpy.org/neps/nep-0018-array-function-protocol.html}{NEP 18})
    mean \astropyQuantity\ works with almost all \package{numpy} (version 1.17+)
    functions.

    In the remaining gaps, the \package{numpy} and \astropyQuantity\
    interoperability efforts are ongoing. For some functions, such as in
    \href{https://numpy.org/doc/stable/user/basics.rec.html#module-numpy.lib.recfunctions}{\package{numpy}'s
    module} for manipulating structured arrays, compatibility only requires
    extending the existing \package{numpy}-\astropypkg bridge framework.
    Community interest, in the form of a Feature Request (or better yet Pull
    Request) would be sufficient to see this realized. For a few remaining
    functions, discussed further in the
    \href{https://docs.astropy.org/en/stable/known_issues.html#known-deficiencies}{\astropy
    documentation}, the \package{numpy} framework does not yet allow for full
    interoperability with \astropy. A goal of \astropy, and the scientific
    \python community is to enhance and implement the frameworks, allowing
    \astropyQuantity\ to propagate units seamlessly across the whole ecosystem
    of \package{numpy}-like projects.

  \paragraph{\astropypkg on Units}

    Units were also integrated further within \astropypkg. For already-defined
    \astropyModel, for instance imported from another library, \astropymodeling
    can now add units to otherwise unit-less models.
    % TODO! transition words
    A new module –
    \astropycosmologyunits – has been added to the cosmology sub-package for
    defining and collecting cosmological units and equivalencies. The unit
    \texttt{littleh} and equivalency
    \astropyapidoc{cosmology.units.with_H0}{with\_H0} are deprecated from the
    main \astropyunits sub-package and moved to \astropycosmologyunits. A new
    unit, \texttt{redshift}, is added for tracking factors of cosmological
    redshift. To correctly use redshift units in dimensional analysis,
    \texttt{redshift} is
    \astropyapidoc{cosmology.units.dimensionless_redshift}{treated as
    dimensionless} by default. To convert between redshift and other
    cosmological distance measures, e.g. CMB temperature or comoving distance,
    the equivalency
    \astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
    is also added. This equivalency is actually a composite of other
    equivalencies, which may be used separately.

  \paragraph{Astropy Fits in Time}
    % IN PROGRESS (ADAPTING FROM THE WHATSNEW).

    The FITS Time standard paper defines the formats and keywords used to
    represent timing information in FITS files. The Astropy FITS package now
    provides support for reading and writing native Time table columns using
    this format. This is described within the FITS unified I/O interface and
    examples of usage can be found in the Astropy native objects (mixin columns)
    section.

@@

  % MHvK: maybe do this before the structure? Change in structure
  % partially driven by this?
    The imrovements to \astropy's I/O architecture are mirrored in the increased support for I/O options.

@@

    First, the class-based architecture allows for the creation of custom
    registries. An application of this is in \astropycosmology, which has two
    different \textit{kinds} of registries for \astropyCosmology: one for
    reading and writing files, and another for converting between python
    objects. Second, \textit{any} kind of object is now compatible with
    \astropyiosubpkg{registry}. Using \astropyiosubpkg{registry} for custom
    classes means a user does not have to create a whole new I/O system and can
    offer a unified and familiar API.

@@

  An important focus of \astropy has been to increase the precision and accuracy
  of its functionality, while, where appropriate, making sure results obtained
  using older versions of \astropypkg are reproducible. Within each section we
  include in our descriptions how old results may be reproduced. The primary
  means are with configuration file options, and settings on runtime
  configuration objects called \astropyScienceState, described further in the
  documentation. Notable improvements to \astropypkg have been made to
  \astropytime, \astropyconstants, \astropycoordinates, and \astropycosmology,
  and we describe each in turn.

  \paragraph{Time}
    In astronomy, time accuracy down to the (nano)second is frequently
    important. Whether for planning observations, crunching pulsar or VLBI data,
    missing seconds meaningfully impact results. Before, when using \astropyTime
    in \astropytime leap seconds had to be manually applied; now leap seconds
    are applied automatically. Moreover, \astropypkg updates internal time
    data-files to ensure that the correct leap second adjustment is always used.
    For reproducibility, e.g. with an old code, \astropyLeapSeconds may also be
    manually applied.
